---
title: "Lab 1"
author: "Wiktor Soral"
date: "February 21 2017"
output: ioslides_presentation
---

##Outline

 - Course aims and structure
 - How to pass the course?
 - Data analysis workflow
 - Data analysis tools
 - Tidy data
 - SPSS - recap
 
##Course aims and structure
  
  - The main goal of this course is to familiarize students with practical knowledge and theoretical background of more advanced data analysis methods.
  - We will aim to the level where you will be able to perform by yourself analyses required for your empirical paper, master thesis, and papers in professional psychological journals.
  - Each lab will begin with short introduction and discussion of the topic, next you will use your knowledge to analyse real word datasets.
  - Home assignments will mostly consist of writing short empirical reports, when you will practice your knowledge even further.
  - To preserve your knowledge you will write 2 exams (midterm and final)

##How to pass the course?

  - Students must respect the principles of academic integrity
  - Attendance: 2 miss classes without excuse, 2 more classes in case of excuse (additional work), more than 4 will not pass
  - Home assignments
  - Midterm and final exam
  - 30% \* (home assignments) + 30% \* (midterm score) + 40% \* (final exam score) = total score
  - Total score and both exam scores should be at least at the 50% level to pass the course.
  - Grading scale: 95%+ = 5! 90-94% = 5 80-89% = 4.5 70-79% = 4 60-69% = 3.5 50-59% 59% = 3 below 50% = 2 (fail)

##Data analysis workflow

  - Collecting your data, ie. running a study
  - Importing your data to stat. software and checking its integrity
  - Assessing whether data are adequate to answer the research question
  - Making necessary transformations, computations, filtering, etc.
  - Building a statistical model that will account for the data
  - Diagnosing problems with the model and the data
  - Assessing whether statistical model fits the data
  - If model accepted: Drawing conclusions from the model -- if not repeat 3 last steps (but with different model)
  - Decide whether to collect more data

##Data analysis tools

  - Graphical user interface, non-free: SPSS, Statistica, Stata, Excel
  - Graphical user interface, free: JASP, PSPP
  
  - Command line interface, non-free: Stata, SAS, S+, MATLAB
  - Command line interface, free: R, some Python packages, Julia

##Tidy data 1

  - You have collected your questionnaires on gender, age, and self-esteem 
  - How should you put them together to work
  
| gender  |count |
|-------- |------|
|  female | 44   |
|  male   | 23   |

##Tidy data 1

  - You have collected your questionnaires on gender, age, and self-esteem 
  - How should you put them together to work
  
|name    |  gender,age,SE |
|--------|----------------|
| Dan    |  male,69,3.6   |
| Brian  |  male,64,3.7   |
| Sue    |  female,65,3.9 |
| Hazel  |  female,68,3.8 |


##Tidy data 2

  - Each variable forms a column
  - Each observation forms a row
  - Each type of observational unit forms a table
  -> Each cell consists of 1 value from 1 subject
  
|name    | gender  | age  | self-esteem (1-4)|
|--------|-------- |------|------------------|
| Dan    |  male   | 69   |  3.6             |
| Brian  |  male   | 64   |  3.7             |
| Sue    |  female | 65   |  3.9             |
| Hazel  |  female | 68   |  3.8             |

##Tidy data 3

 - Ordering of columns - first fixed values, then random values
 - Missing values are identified
 - Value meaning is indicated (where possible)
 - Scale type is identified

##Assumptions 1

  - Normally distributed data
  
```{r echo=FALSE}
hist(rnorm(1000),freq = F, col = "lightblue", main ="", xlab='')
curve(dnorm(x),add = T, col="red", lw=2)
```

##Assumptions 1

  - Non-normally distributed data
  
```{r echo=FALSE}
hist(rpois(1000, lambda = 3),freq = F, breaks=10, col = "lightblue", main ="", xlab='')
curve(dnorm(x, mean = 3, sd = sqrt(3)),add = T, col="red", lw=2)
```

  
##Assumptions 2

  - Homogeneity of variance
  
```{r echo=FALSE}
boxplot(rnorm(500, 3, 2), rnorm(500, 5, 2), col = "lightblue")
```

##Assumptions 2

  - Heterogeneity of variance
  
```{r echo=FALSE}
boxplot(rnorm(500, 3, 2), rnorm(500, 5, 4), col = "lightblue")
```

##Assumptions 3

  - Interval data
  
```{r echo=FALSE}
catego = sample(1:10, 400, replace= T, c(0.02, 0.03,0.10,0.15, 0.20, 0.20,0.15, 0.10,0.03, 0.02))
barplot(prop.table(table(catego)), ylim =c(0, 0.3),col='lightblue')
```


##Assumptions 3

  - Non-interval data
  
```{r echo=FALSE}
catego = sample(c('a','b','c','d'), 200, replace= T, prob = c(0.45, 0.25, 0.20, 0.1))
barplot(prop.table(table(catego)), ylim =c(0, 0.5),col='lightblue')
```

##Assumptions 4

  - Independence
  
```{r echo=FALSE}
x <- rnorm(1000)
acf(x, main = "")
```


##Assumptions 4

  - Non-independence
  
```{r echo=FALSE}
x <- filter(rnorm(1000), filter=rep(1,7), circular=TRUE)
acf(x, main = "")
```

##SPSS - recap

 1. Selecting a subset of data, splitting data
 2. Computing new values and recoding data
 3. Making simple graphs
 4. Checking normality with histogram and QQ-plot
 5. Checking homogeneity of variance with boxplots
 6. Checking independence of observations with auto-correlation plots
 7. Detecting outliers

